Measuring Interaction Design

### 2015-09-24 paper.tex
completed "how it works" section

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\newcommand{\mytodo}[1]{\todo[inline]{#1}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\migtool}{MIG-Tool\ }
\usepackage{soul}
\usepackage{paralist}
\usepackage{listings}

@@

\keywords{Experimental; Evaluation; Statecharts: UML; UML-IDEA; Testing.}

@@

In many cases \emph{statecharts} are used (a generalization of finite
state automata, with an expressive language that includes hierarchic
levels of abstraction, concurrent regions, states and pseudostates, an
extended state notion based on an arbitrary computational model).
Harel, the inventor of statecharts, gives an interesting retrospective
view of how they were invented, back in early eighties, and why they
were appealing also to non experts~\cite{harel09}.  For the sake of
brevity, we refrain from describing them here in detail; the
interested reader is referred to the UML standard~\cite{uml241} or
some of the textbooks that deal with them, like~\cite{samek09}.
%
Horrocks showed how statecharts can be  used to model and
specify the dynamics of typical desktop UIs~\cite{horrocks99}. While
providing many interesting insights on how and why one should use
statecharts to do so, this nice work does not address how such a
specification could be \emph{automatically} processed.
%
This idea was later on greatly expanded by
Thimbleby~\cite{thimbleby07} , which discusses the use of statecharts
as meta-models for user interfaces; they are seen as crucial
representations that allow a designer to fully appreciate how devices
behave. The overall claim is that ``If you don't understand the logic
conveyed by a statechart model of a user interface, then you don't
understand the behavior of the user interface''.

@@

Markov models, \ie\ directed graphs where edges leaving a vertex are
associated to a probability distribution, were used
in~\cite{thimbleby2001markov}, as a means to perform usability
analysis as early as possible, even before building prototypes of the
UI. Vertices represent states of the UI and edges correspond to user
actions (such as pressing a push-down button).  Probabilities can be
used as a model of the skills or knowledge of the user: equiprobable
actions correspond to a knowledge-free user, whereas when some actions
have a very low probability it means that for that user the action is
unlikely to be executed (perhaps because little known). Simple
mathematical operations on the transition matrix of the model give the
probability that after $n$ steps from a given initial state, the UI is
in a given state. With Markov models, by manipulating probabilities,
the designer can plot the number of required steps as a function of
how close the probabilities are to the designer's ``perfect''
knowledge.  Examples discussed in the paper cover several devices,
ranging from a simple torch (with 4 states), a microwave cooker (6
states), a mobile phone (152 states), but all being push-down devices
with a fixed set of buttons. This is obviously not the case for UIs of
information systems, where some buttons may or may not be ``painted''
on the screen. This makes it more difficult to specify the transition
matrix. Another limitation of this approach is that probabilities are
assumed to be constant, corresponding to users that are not learning
while interacting with the system.
%
Our approach is based on statecharts, a language that in practice is
much more concise than Markov models, making it easier to specify the
UI behavior, especially in cases where the set of buttons change over
time. While our examples do not make use of probabilities, this is
very easy to cope with (see the Discussion at the end of the paper for
some of the benefits that doing this could bring). Similarly
to~\cite{thimbleby2001markov}, our approach could be used when
conservative results that do not rely on psychological assumptions are
sought.

centrality measures

@@

@@

\section{Generation of interaction paths}
\label{sec:traces}

\begin{quote}
  here we describe how we transform state machines into graphs, how we
  specify scenarios, how the system computes traces, how metrics are
  computed. 
\end{quote}

outline:
- modeling, statecharts
- flattening, ixgraph generation
- scenario def
- generation of traces
- analysis

modeling:
- statecharts details

flattening state machine \cite{briand2005:test:uml}
-

Notice that a scenario specifies only the desired occurrences of user
actions, not the necessary ones. For example, the model specifies that
in order to view a conversation while reading another one, one has to
\texttt{goBack} to the inbox first. It is the task of \migtool to
unfold cyles in the graph and compute all the  paths that connect the
desired user actions.

Actions specified in a scenario are called \texttt{bridges}, because
they represent one or more edges in the graph that have to be followed
when carrying out the

issues dealing with complexity of searching paths

@@

\migtool is part of a suite of model-based tools for designing,
analysing and testing user interfaces developed by the company
ANONYMIZED. Development of a web-based user interface of \migtool is
underway; the company plans to make it freely available for research
purposes.

>>>>

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\newcommand{\mytodo}[1]{\todo[inline]{#1}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\migtool}{MIGtool\ }
\usepackage{soul}
\usepackage{paralist}
\usepackage{listings}

@@

\keywords{Experimental; Evaluation; Statecharts: UML; Metrics; Testing.}

@@

In many cases \emph{statecharts} are used, a generalization of finite
state automata, which is a more expressive language.
%
Horrocks showed how statecharts can be  used to model and
specify the dynamics of typical desktop UIs~\cite{horrocks99}. While
providing many interesting insights on how and why one should use
statecharts to do so, this nice work does not address how such a
specification could be \emph{automatically} processed.
%
This idea was later on expanded by
Thimbleby~\cite{thimbleby07} , which discusses the use of statecharts
as meta-models for user interfaces; they are seen as crucial
representations that allow a designer to fully appreciate how devices
behave. The overall claim is that ``If you don't understand the logic
conveyed by a statechart model of a user interface, then you don't
understand the behavior of the user interface''.

@@

Markov models, \ie\ directed graphs where edges leaving a vertex are
associated to a probability distribution, were used
in~\cite{thimbleby2001markov}, as a means to perform usability
analysis as early as possible, even before building prototypes of the
UI. Vertices represent states of the UI and edges correspond to user
actions (such as pressing a push-down button).  Probabilities can be
used as a model of the skills or knowledge of the user: equiprobable
actions correspond to a knowledge-free user, whereas when some actions
have a very low probability it means that for that user the action is
unlikely to be executed (perhaps because little known). Simple
mathematical operations on the transition matrix of the model give the
probability that after $n$ steps from a given initial state, the UI is
in a given state. With Markov models, by manipulating probabilities,
the designer can plot the number of required steps as a function of
how close the probabilities are to the designer's ``perfect''
knowledge.  Examples discussed in the paper cover several devices,
ranging from a simple torch (with 4 states), a microwave cooker (6
states), a mobile phone (152 states), but all being push-down devices
with a fixed set of buttons. This is obviously not the case for UIs of
information systems, where some buttons may or may not be ``painted''
on the screen. This makes it more difficult to specify the transition
matrix. Another limitation of this approach is that probabilities are
assumed to be constant, corresponding to users that are not learning
while interacting with the system.
%
Our approach is based on statecharts, a language that in practice is
more expressive than Markov models, making it easier to specify the
UI behavior, especially in cases where the set of buttons change over
time. While our examples do not make use of probabilities, this is
very easy to cope with (see the Discussion at the end of the paper for
some of the benefits that doing this could bring). Similarly
to~\cite{thimbleby2001markov}, our approach could be used when
conservative results that do not rely on psychological assumptions are
sought.

@@

vedere \cite{hilbert2000}
\mytodo{However, because user interface
  events are typically voluminos and rich in detail, automated support
  is generally required to extract information at a level of
  abstraction that is useful to investigators interested in analyzing
  application usage or evaluating usability. This survey examines
  computer-aided techniques used by HCI practitioners and researchers
  to extract usability-related information from user interface
  events. A framework is presented to help HCI practitioners and
  researchers categorize and compare the approaches that have been, or
  might fruitfully be, applied to this problem.

  They - mention levels of abstractions in user events; what we deal
  with here are ``abstract interaction level'' events - they mention
  the problem of detecting sequences from event streams, which in a
  sense is the converse of what we do - they discuss partial matches
  between event streams and ``target'' sequences; we do the converse:
  generation of event streams with detours from target sequences.  }

@@

\section{Generation of interaction traces}
\label{sec:traces}

Interaction traces are paths (\ie, sequences of connected states
of the model) that users can follow to perform a given scenario.  The
generation process includes the following steps:
\begin{inparaenum}[(1)]
\item automatic flattening of the statechart model;
\item manual definition of the interaction scenario;
\item automatic generation of execution traces; and
\item manual interactive analysis of results.
\end{inparaenum}

\subsection{Processing models }

\migtool\ takes as input UML statecharts, which are a
generalization of finite state automata (FSA), with an expressive language
that includes hierarchic levels of abstraction, concurrent regions,
states and pseudostates, an extended state notion based on an
arbitrary computational model.  Harel, the inventor of statecharts,
gives an interesting retrospective view of how they were invented,
back in early eighties, and why they were appealing also to non
experts~\cite{harel09}.  For the sake of brevity, we refrain from
describing them here in detail; the interested reader is referred to
the UML standard~\cite{uml25} or some of the textbooks that deal with
them, like~\cite{samek09}.

Using statecharts, behavior of UIs can be represented by associating
states to screens and particular configuration of widgets, and
transitions to actions performed by users or by the system itself.

Because statechart models take advantage of abstraction features and a
rich set of connecting pseudostates, they are not suitable to be
directly processed to compute metrics. For this reason, \migtool first
\emph{flattens} the model. Flattening is a process often used when
statecharts have to be automatically
processed~\cite{briand2005:test:uml}, and it means to produce a FSA
that is behaviorally equivalent to the original statechart, where
hierarchy between states and concurrent regions are removed. In most
cases this leads to an exponential number of states and transitions,
but because the process is completely automatic and in our case there
is no need to manually inspect the resulting FSA, this aspect is in
many cases not relevant.  Flattening statecharts is an often used
process when they have to be automatically
processed~\cite{briand2005:test:uml}.  \migtool\ produces an XML
representation (graphML) of the resulting FSA, the \emph{interaction
  graph}. It is a directed multigraph, potentially with cycles and
loops, with edges that are labeled with the name of the corresponding
action\footnote{A directed multigraph is a graph such that there is at
  least a pair of vertices that are connected by 2 or more edges that
  have the same direction. Cycles are paths that include 2 or more
  occurrences of the same edge. Loops are edges that start and end on
  the same vertex.}. In the following we will be referring to the
interaction graph, and will be using as synonymous the terms ``state''
and ``vertex'', and ``action'' and ``edge''.

\subsection{Defining usage scenarios}

Because in all but the most trivial interaction graphs there are
cycles, the set of possible interaction traces is usually
infinite. For this reason, scenarios need to be defined as constraints
on the possible execution traces that can be generated.  Users of
\migtool\ define interaction scenarios by specifying key steps (called
\emph{bridge sets}) that users are expected to go through; we call
this process \emph{grounding usage scenarios on models}. Each bridge
set is specified by selecting a subset of the edges of the interaction
graph. Often a scenario might require also optional steps. In such a
case, the specification includes an initial state and a sequence of
bridge sets. For example, to specify a scenario for replying to an
email message, one could select all the edges associated to the action
``reply'' (bridge set 1), followed by edges labeled with ``typeBody''
(bridge set 2), followed by edges labeled with ``send'' (bridge set
3). A well formed bridge set would consist of 1 or more edges (a
bridge set with 0 edges would make the scenario unviable). In this way
scenarios with cycles can be formulated (\eg, reply to two
messages). Each pair of consecutive bridge sets makes a \emph{stage}
of the scenario.
%
To cope with multi edges, the interaction graph is \emph{simplified}:
all edges between  pairs of vertices are merged into a single one,
whose label includes the original ones.

\subsection{Searching traces}

Quite expressive languages can be conceived for grounding scenarios
(\eg, languages based on regular expressions). But such expressivity
bonus needs to be balanced with computational tractability: even
models with 50 states might lead to interaction graphs with several
hundred states and several thousand edges, leading to an enormous
number of possible paths to filter even for scenarios with just a few
stages.

To cope with this we implemented a path searching algorithm that
processes each of the stages sequentially, starting from the initial
initial state. Given a bridge set $B_i$, the algorithm does a
breadth-first search of all the geodesic paths\footnote{``Geodesic''
  is synonymous with ``shortest''.}  that connect each of the ending
vertices of edges in the bridge set $B_i$ with some of the starting
vertices of edges in bridge set $B_{i+1}$. If some of the starting
vertices cannot be reached, then the bridge is dropped. \migtool\
creates a new graph from the geodesic paths found for each stage, and
then joins these graphs so that geodesic paths found for stage $i$ are
joined with those of ${i+1}$. These global paths, connecting the
initial state and the remaining bridges of the last bridge set of the
scenario are called \emph{execution traces}.

Notice that a scenario specifies only the desired occurrences of
actions, not all the necessary ones. For example, if the model
prescribes that in order to \texttt{view(message)} while reading
another one, one has to \texttt{goBack} to the inbox first, a scenario
specifying two consecutive \texttt{view(message)} would lead to
geodesic paths that include also the \texttt{goBack} action, even if
it is not specified in the scenario. It is the task of \migtool to
unfold cyles in the graph and search all the geodesic paths that
connect the desired user actions.

\subsection{Detour traces}
The algorithm described so far finds (some of) the global paths
connecting the initial state to one or more bridges for each of the
specified stages. The globally shortest path is identified, together
with other viable alternatives. These traces are called
\emph{detour order 0} traces (and states).

For each stage, up to a maximum detour order $H$, the search algorithm
creates traces with detour order $k+1$ by collecting states $D_k$
with order $0$ up to $k$, and by identifying the neighbours ($N_k$)
of $D_k$ (which are the states not included in $D_k$ that are
connected through an edge to some state in $D_k$). Edges connecting
$D_k$ with $N_k$ represent deviations that users might follow, and
geodesic paths between $N_k$ and $D_k$ constitute recovery paths that
users might follow to complete the scenario from states in
$N_k$. These deviations and recovery paths are joined and constitute
the traces of order $k+1$.  It could happen that
for some state in $N_k$ there is no path leading to any vertex in
$D_k$: in such a case the deviation leads to a dead end that prevents
the user to complete the scenario.

In general, \migtool is used to process a model against a scenario
and to generate execution traces of detour order = 0,
..., H. The time complexity of the search algorithm is
$\mathcal{O}(KM(E+V)))$, where there are $K$ bridge sets, the mean
size of them is $M$, the interaction graph has $V$ vertices and $E$
edges. Therefore it scales pretty well with the size of the
interaction graph and/or complexity of the scenarios. In practice for
interaction graphs consisting of about 10000 edges and a dozen of
bridge sets, on a low-cost PC it takes about 15 seconds to generate
traces of order 0 to 3.

@@

\migtool is part of a suite of model-based tools for designing,
analysing and testing user interfaces developed by the company
ANONYMIZED. \migtool is implemented partly in Java (model processing)
and partly in R/iGraph. Development of a web-based UI of \migtool is
underway; \migtool will be made freely available for research
purposes.
